Supplementary Information

### 2021-06-10 doc/manuscript/ForestGEO_treerings.Rmd
google docs changes- discussion

Tree rings provide an invaluable long-term record for understanding how climate and other drivers shape tree growth and forest productivity.
However, conventional tree-ring analysis methods were not designed to simultaneously test effects of climate, tree size, and other drivers on individual growth.
This has limited the potential to use tree rings to understand forest productivity, its climate sensitivity, and its global change responses.
Here, we develop and apply a new method to simultaneously model non-linear effects of primary climate drivers, reconstructed tree diameter (DBH), and calendar year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
We analyze data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that precipitation, temperature, DBH, and calendar year have additively, and often interactively, influenced annual growth over the past 120 years.
Growth responses were predominantly positive to precipitation (usually over $\ge$ 3-month seasonal windows) and negative to temperature (usually over $\le$ 3-month seasonal windows), with both included in 78% of top models, and with non-linear responses prevalent (63% of relationships).
Climate sensitivity commonly varied with DBH (44% of cases tested). 
Trends in ring width at small DBH were linked to the light environment under which trees established, but basal area or biomass increments consistently peaked at intermediate DBH and declined thereafter.
Accounting for climate and DBH, growth rate declined over time for 92% of species in secondary or disturbed stands, whereas growth trends were mixed in older forests.
These trends were largely attributable to stand dynamics as cohorts and stands age, which remain challenging to disentangle from global change drivers.
By providing a parsimonious approach for characterizing multiple interacting drivers of tree growth, our method reveals a more complete picture of the factors influencing growth than has previously been possible.

@@

To realistically estimate forest woody productivity, it is necessary to measure or model the growth rate of individual trees within a stand based on the primary biotic and abiotic drivers.
Specifically, what is needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter breast height, DBH), and other environmental drivers, all of which may be best described by nonlinear functions [@muller-landau_testing_2006; @rollinson_climate_2021].
While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied in tree-ring analysis [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously account for climate, DBH, and calendar year (a proxy for slowly changing environmental drivers).
In the following paragraphs, we outline major hypotheses regarding the influence of these factors on tree growth that can best be addressed using a multifactorial, non-linear approach to tree-ring analysis (Table `r hypotheses`).

@@

Tree DBH scales predictably with numerous traits affecting tree growth (e.g., height, crown size and position, root mass, hydraulic architecture) and therefore is itself linked to growth [e.g., @anderson-teixeira_sizerelated_2015].
Yet, there remain inconsistencies across disciplines (dendrochronology, forest ecology) as to what is considered a typical growth pattern. 
Dendrochronological records from trees that established in high-light environments commonly show a pattern in which radial stem growth increments (ring width, RW) are initially large and decline with DBH [@fritts_tree_1976], whereas censuses of globally distributed forests indicate that diameter increments most commonly increase with DBH [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
We hypothesize that this discrepancy is primarily a distinction between trees that establish in the open versus in the understory [Table `r hypotheses`, @lorimer_estimating_1988].
Building upon observed ontogenetic patterns in RW, dendrochronology studies often adopt a null hypothesis that the annual basal area increment (BAI) fluctuates around a constant mean after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
Finally, there is debate as to whether the aboveground biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks at intermediate DBH and then plateaus or declines as trees divert carbon to other functions such as reproduction and respiration [@thomas_agerelated_2011; @sheil_does_2017; @west_increasing_2020]. 
Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
Discerning these ontogenetic growth trends is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.

@@

We analyzed tree-ring data, most of which was collected for earlier studies (see references in Table `r sites`), from 10 sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and conifer, and boreal conifer (Tables `r sites`, `r site_details`, `r species_table`).
Nine of these sites (exception: LT) were co-located with large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
Trees were cored within the ForestGEO plots (n=5 sites) and/or nearby within similar forest types (n=5 sites), following a variety of sampling protocols designed to meet the objectives of the original studies (Tables `r site_details`, `r sampling_details_table`). 
In using this diversity of data sources, we ensured that our approach could handle challenges presented by varying methodologies and forest types.

@@

Our finding that BAI and $\Delta$AGB generally saturate or decline with increasing DBH (Table `r hypotheses`, Fig. `r size`), contrasts with findings of cross-sectional analyses of forest census data showing that $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018].
In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021].
Declines in BAI and $\Delta$AGB at larger DBH are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
Notably, inclusion of year in the GLS models tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs. `r GLS_SCBI`, `r GLS_SCBI_year`), suggesting that some of the declines at large DBH (Fig. `r size`) are more properly attributed to recent environmental changes than to large DBH.

@@

Our analytical framework reconstructs growth changes in a sampled tree population over time while accounting for climate, DBH, and persistent growth differences among individuals (Fig. `r schematic`), thereby addressing some important challenges to obtaining unbiased estimates of growth trends attributable to non-climatic environmental drivers.
First, we correct for changes in climate that may drive directional growth trends.
For example, dramatic growth declines at LT, driven by a strong regional warming and drying trend [@touchan_millennial_2011; @williams_temperature_2013], are in part factored out by accounting for the primary climate drivers, such that a significant decline over time was detected for only one of the two species (Fig. `r year_fig`).
Second, we show that growth rate -- by any metric -- varies nonlinearly with DBH and with patterns dependent upon the species and environmental context (Fig. `r size`), reinforcing the concept that growth trend analyses should incorporate cross-sectional analyses to correct for growth ontogeny [e.g., through regional curve standardization, @peters_detecting_2015].
Our method does this, differing from the conceptually parallel method of regional curve standardization in that we standardize relative to DBH rather than age, correct for any trends in the most influential climate drivers, and include random effects of tree to account for persistent growth differences among individuals.
The latter addresses a third important challenge, which is accounting for persistent growth differences among individuals that can bias estimated growth trends in positive or negative directions [@brienen_detecting_2012;@nehrbass-ahles_influence_2014; @groenendijk_no_2015; @brienen_tree_2017; @vandersleen_trends_2017].
For instance, older trees, which provide the only records available for the earliest decades, are competitive winners that probably had above-average growth rates [@aubry-kientz_joint_2015], which would upwardly bias average growth rate estimates for early decades [@groenendijk_no_2015].
By including a random effect of tree, our approach avoids the the most severe potential biases associated with persistent growth differences across individuals [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013], yet observed trends nevertheless represent only the sampled population of trees, as opposed to tree populations at all points throughout the time frame analyzed.
Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., indirect effects of climate change, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).

In all seven sites dominated by trees less than 100 years old, growth trends were universally negative, when significant (Table `r hypotheses`, Fig. `r year_fig`). 
These trends are probably attributable primarily to stand dynamics as cohorts and stands develop over time, and are fairly typical of mixed-species stands that experience vertical stratification [@oliver_forest_1990]. 
For species exhibiting a pulse of recruitment in the past followed by little subsequent recruitment (e.g., *Acer rubrum* and *Betula alleghaniensis* at HF), persistent differences in growth rates among individuals could produce a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017; see also @vandersleen_trends_2017].
Particularly in secondary stands where many of the sampled species recruited in pulses that were followed by low recruitment (e.g., SCBI, HF; Appendix `r site_details_appendix`), growth declines are consistent with the tendency for faster tree growth during early stand development [@lorimer_estimating_1988; @lorimer_methodology_1989; @oliver_forest_1990], and with increasing competition and declining woody productivity as young stands mature [e.g., @goulden_patterns_2011; @pregitzer_carbon_2004; @west_increasing_2020].
Abiotic drivers likely also play a role at some of these sites.
At Scotty Creek, in northern Canada, rapid warming is thawing permafrost and altering hydrologic conditions [@baltzer_forests_2014], resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020]; we attribute pronounced negative growth trends to a combination of successional declines and indirect climatic stress.

>>>>

Tree rings provide an invaluable long-term record for understanding how climate and other drivers shape tree growth and forest productivity.
However, conventional tree-ring analysis methods were not designed to simultaneously test effects of climate, tree size, and other drivers on individual growth.
This has limited the potential to use tree rings to understand forest productivity, its climate sensitivity, and its global change responses.
Here, we develop and apply a new method to simultaneously model non-linear effects of primary climate drivers, reconstructed tree diameter (DBH), and calendar year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
We analyze data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that precipitation, temperature, DBH, and calendar year have additively, and often interactively, influenced annual growth over the past 120 years.
Growth responses were predominantly positive to precipitation (usually over $\ge$ 3-month seasonal windows) and negative to temperature (usually over $\le$ 3-month seasonal windows), with both included in 78% of top models, and with non-linear responses prevalent (63% of relationships).
Climate sensitivity commonly varied with DBH (44% of cases tested). 
Trends in ring width at small DBH were linked to the light environment under which trees established, but basal area or biomass increments consistently reached a maximum at intermediate DBH.
Accounting for climate and DBH, growth rate declined over time for 92% of species in secondary or disturbed stands, whereas growth trends were mixed in older forests.
These trends were largely attributable to stand dynamics as cohorts and stands age, which remain challenging to disentangle from global change drivers.
By providing a parsimonious approach for characterizing multiple interacting drivers of tree growth, our method reveals a more complete picture of the factors influencing growth than has previously been possible.

@@

To realistically estimate forest woody productivity, it is necessary to measure or model the growth rate of individual trees within a stand based on the primary biotic and abiotic drivers.
Specifically, what is needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter at breast height, DBH), and other environmental drivers, all of which may be best described by nonlinear functions [@muller-landau_testing_2006; @rollinson_climate_2021].
While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied in tree-ring analysis [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously account for climate, DBH, and calendar year (a proxy for slowly changing environmental drivers).
In the following paragraphs, we outline major hypotheses regarding the influence of these factors on tree growth that can best be addressed using a multifactorial, non-linear approach to tree-ring analysis (Table `r hypotheses`).

@@

Tree DBH scales predictably with numerous traits affecting tree growth (e.g., height, crown size and position, root mass, hydraulic architecture) and therefore is itself linked to growth [e.g., @anderson-teixeira_sizerelated_2015].
Yet, there remain inconsistencies across disciplines (dendrochronology, forest ecology) as to what is considered a typical growth pattern. 
Dendrochronological records from trees that established in high-light environments commonly show a pattern in which radial stem growth increments (ring width, RW) are initially large and decline with DBH [@fritts_tree_1976], whereas censuses of globally distributed forests indicate that diameter increments most commonly increase with DBH [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
We hypothesize that this discrepancy is primarily a distinction between trees that establish in the open versus in the understory [Table `r hypotheses`, @lorimer_estimating_1988].
Building upon observed ontogenetic patterns in RW, dendrochronology studies often adopt a null hypothesis that the annual basal area increment (BAI) fluctuates around a constant mean after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
Finally, there is debate as to whether the aboveground biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016; @sillett_increasing_2010] or peaks at intermediate DBH and then plateaus or declines as trees divert carbon to other functions such as reproduction and respiration [@thomas_agerelated_2011; @sheil_does_2017; @west_increasing_2020]. 
Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
Discerning these ontogenetic growth trends is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.

@@

We analyzed tree-ring data, most of which was collected for earlier studies (see references in Table `r sites`), from 10 sites ranging from 9.15$^\circ$ to 61.30$^\circ$N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and conifer, and boreal conifer (Tables `r sites`, `r site_details`, `r species_table`).
Nine of these sites (exception: LT) were co-located with large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
Trees were cored within the ForestGEO plots (n=5 sites) and/or nearby within similar forest types (n=5 sites), following a variety of sampling protocols designed to meet the objectives of the original studies (Tables `r site_details`, `r sampling_details_table`). 
In using this diversity of data sources, we ensured that our approach could handle challenges presented by varying methodologies and forest types.

@@

Our finding that BAI and $\Delta$AGB generally saturate or decline with increasing DBH (Table `r hypotheses`, Fig. `r size`), contrasts with findings of cross-sectional analyses of forest census data showing that $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018].
In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021].
Declines in BAI and $\Delta$AGB at larger DBH are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
Apparent declines in $\Delta$AGB at large DBH (or old age) may also be driven by shifts towards proportionally greater wood production within the crown [e.g., branch production, @sillett_increasing_2010b; @sillett_comparative_2021] that are not adequately captured by biomass allometries based on DBH and sometimes height [@goodman_importance_2014b; @disney_new_2020].
Notably, inclusion of year in the GLS models tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs. `r GLS_SCBI`, `r GLS_SCBI_year`), suggesting that some of the declines at large DBH (Fig. `r size`) are more properly attributed to recent environmental changes than to large DBH.

@@

Our analytical framework reconstructs growth changes in a sampled tree population over time while accounting for climate, DBH, and persistent growth differences among individuals (Fig. `r schematic`), thereby addressing some important challenges to obtaining unbiased estimates of growth trends attributable to non-climatic environmental drivers.
First, we correct for changes in climate that may drive directional growth trends.
For example, dramatic growth declines at LT, driven by a strong regional warming and drying trend [@touchan_millennial_2011; @williams_temperature_2013], are in part factored out by accounting for the primary climate drivers, such that a significant decline over time was detected for only one of the two species (Fig. `r year_fig`).
Second, we show that growth rate -- by any metric -- varies nonlinearly with DBH and with patterns dependent upon the species and environmental context (Fig. `r size`), reinforcing the concept that growth trend analyses should incorporate cross-sectional analyses to correct for growth ontogeny [e.g., through regional curve standardization, @peters_detecting_2015].
Our method does this, differing from the conceptually parallel method of regional curve standardization in that we standardize relative to DBH rather than age, correct for any trends in the most influential climate drivers, and include random effects of tree to account for persistent growth differences among individuals.
The latter addresses a third important challenge, as those growth differences among individuals can bias estimated growth trends in positive or negative directions [@brienen_detecting_2012;@nehrbass-ahles_influence_2014; @groenendijk_no_2015; @brienen_tree_2017; @vandersleen_trends_2017].
For instance, older trees, which provide the only records available for the earliest decades, are competitive winners that may have had above-average growth rates [@aubry-kientz_joint_2015], which would upwardly bias average growth rate estimates for early decades [@groenendijk_no_2015].
By including a random effect of tree, our approach likely reduces the most severe potential biases associated with persistent growth differences across individuals [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013], yet observed trends nevertheless represent only the sampled population of trees, as opposed to tree populations at all points throughout the time frame analyzed.
Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., indirect effects of climate change, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).

In all seven sites dominated by trees less than 100 years old, growth trends were universally negative, when significant (Table `r hypotheses`, Fig. `r year_fig`). 
While attribution of these trends is difficult, it is likely that some trends are caused by stand dynamics as cohorts and stands develop over time [@oliver_forest_1990]. 
For species exhibiting a pulse of recruitment in the past followed by little subsequent recruitment (e.g., *Acer rubrum* and *Betula alleghaniensis* at HF), persistent differences in growth rates among individuals could produce a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017; see also @vandersleen_trends_2017].
Particularly in secondary stands where many of the sampled species recruited in pulses that were followed by low recruitment (e.g., SCBI, HF; Appendix `r site_details_appendix`), growth declines are consistent with the tendency for faster tree growth during early stand development [@lorimer_estimating_1988; @lorimer_methodology_1989; @oliver_forest_1990], and with increasing competition and declining woody productivity as young stands mature [e.g., @goulden_patterns_2011; @pregitzer_carbon_2004; @west_increasing_2020].
Abiotic drivers likely also play a role at some of these sites.
At Scotty Creek, in northern Canada, rapid warming is thawing permafrost and altering hydrologic conditions [@baltzer_forests_2014], resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020]; we attribute pronounced negative growth trends to a combination of successional declines and indirect climatic stress.
